Reducing Linearizability of Priority Queues and More Data Structures to State Reachability

### 2017-07-13 checking-executions-arxiv.tex
arxiv

%TODO REPHRASE THIS
%While checking linearizability of a given execution is known to be NP-complete in general~\cite{journals/siamcomp/GibbonsK97} we show in this section that it is polynomial-time for data-differentiated executions of the priority queue.
We define a recursive procedure for checking linearizability of an execution w.r.t. $\seqPQ$.
To ease the exposition, Section~\ref{ssec:seq_exec} introduces a recursive procedure for checking whether a data-differentiated \emph{sequential} execution is admitted by the priority queue which is then extended to the concurrent case in Section~\ref{ssec:conc_exec}.

@@

Checking whether a data-differentiated sequential execution belongs to $\seqPQ$ could be implemented by checking membership into the set of traces of the LTS $PQ$. The recursive procedure $\textit{Check-PQ-Seq}$ outlined in Algorithm~\ref{alg:seq_check} is an alternative to this membership test. Roughly, it selects one or two operations in the input execution, checks whether their return values are correct by ignoring the order between the other operations other than how they are ordered w.r.t. the selected ones, and calls itself recursively on the execution without the selected operations. %{\color {blue} Here we implicitly mix operation $o$ (when $\textit{call}_o(m,a)$, $\textit{ret}_o(m,b)$ in $e$) with $m(a,b)$.}

We explain how the procedure works on the following execution:
{
\begin{align}
\hspace{-5mm}\textit{put}(c,p_2) \cdot \textit{put}(a,p_1) \cdot \textit{rm}(a) \cdot \textit{rm}(c) \cdot \textit{rm}(\textit{empty}) \cdot \textit{put}(d,p_2) \cdot \textit{put}(f,p_3) \cdot \textit{rm}(f) \cdot \textit{put}(b,p_1)\label{eq:ex_rec1}
\end{align}
}
where $p_1$, $p_2$, $p_3$ are priorities such that $p_1 \prec p_2$ and $p_1 \prec p_3$, and $p_2$ and $p_3$ are incomparable. Since the $\textit{rm}(\textit{empty})$ operations %\footnote{An $m(a)$-operation in an execution $e$ is an operation identifier $o$ s.t. $e$ contains the actions $\textit{call}_o(m,a)$, $\textit{ret}_o(m,a)$.}
are read-only (they don't affect the state of the queue), they are selected first. Ensuring that an operation $o=\textit{rm}(\textit{empty})$ is correct boils down to checking that every $\textit{put}(x,p)$ operation before $o$ is matched to a $\textit{rm}(x)$ operation which also occurs before $o$. This is true in this case for $x\in \{a,c\}$. Therefore, the correctness of (\ref{eq:ex_rec1}) reduces to the correctness of
{
\begin{align*}
\textit{put}(c,p_2) \cdot \textit{put}(a,p_1) \cdot \textit{rm}(a) \cdot \textit{rm}(c) \cdot \textit{put}(d,p_2) \cdot \textit{put}(f,p_3) \cdot \textit{rm}(f) \cdot \textit{put}(b,p_1)
\end{align*}
}
When there are no more $\textit{rm}(\textit{empty})$ operations, the procedure selects a $\textit{put}$ operation adding a value with maximal priority which is not removed, and then a pair of $\textit{put}$ and $\textit{rm}$ operations adding and removing the same maximal priority value. For instance, since $p_2$ is a maximal priority, it selects the operation $\textit{put}(d,p_2)$. {This operation is correct if $d$ is the last value with priority $p_2$, and the correctness of (\ref{eq:ex_rec1}) reduces to the correctness of
{
\begin{align*}
\textit{put}(c,p_2) \cdot \textit{put}(a,p_1) \cdot \textit{rm}(a) \cdot \textit{rm}(c) \cdot \textit{put}(f,p_3) \cdot \textit{rm}(f) \cdot \textit{put}(b,p_1)
\end{align*}
}
Since there is no other value of maximal priority which is not removed, the procedure selects a pair of $\textit{put}$/$\textit{rm}$ operations with an argument of maximal priority $p_2$,} for instance, $\textit{put}(c,p_2)$ and $\textit{rm}(c)$. The value returned by $\textit{rm}$ is correct if all the values of priority smaller than $p_2$ added before $\textit{rm}(c)$ are also removed before $\textit{rm}(c)$. In this case, $a$ is the only value of priority smaller than $p_2$ and it satisfies this property. Applying a similar reasoning for all the remaining values, it can be proved that this execution is correct.

\begin{algorithm}[t]
\KwIn {A data-differentiated sequential execution $e$}
\KwOut{$\mathsf{true}$ iff $e\in \seqPQ$}

@@

\If {$\mathsf{Has\text{-}EmptyRemoves}(e)$}
{
    \If {$\exists\ o=\textit{rm}(\textit{empty})\in e$ such that $\mathsf{EmptyRemove\text{-}Seq}(e,o)$ holds}
    {
        \KwRet $\textit{Check-PQ-Seq}(e \setminus o)$\;
    }
    %\Else {\KwRet $\mathsf{false}$\;}
}
\ElseIf{$\mathsf{Has\text{-}UnmatchedMaxPriority}(e)$}
{
    \If {$\exists\ x \in \textit{values}(e)$ such that $\mathsf{UnmatchedMaxPriority\text{-}Seq}(e,x)$ holds}
    {
        \KwRet $\textit{Check-PQ-Seq}(e \setminus x)$\;
    }
    %\Else {\KwRet $\mathsf{false}$\;}
}

\Else
{
    \If {$\exists\ x \in \textit{values}(e)$ such that $\mathsf{MatchedMaxPriority\text{-}Seq}(e,x)$ holds}
    {
        \KwRet $\textit{Check-PQ-Seq}(e \setminus x)$\;
    }
    \Else {\KwRet $\mathsf{false}$\;}
}
\caption{$\textit{Check-PQ-Seq}$}
\label{alg:seq_check}
\end{algorithm}

In formal terms, the operations which are selected depend on the following set of predicates on executions:
{\small
\begin{align*}
\mathsf{Has\text{-}EmptyRemoves}(e)=\mathsf{true} & \mbox{ iff  $e$ contains a $\textit{rm}(\textit{empty})$ operation} \\
\mathsf{Has\text{-}UnmatchedMaxPriority}(e)=\mathsf{true} & \mbox{ iff $p\in \textit{unmatched-priorities}(e)$ for a maximal priority} \\
&\hspace{4mm}\mbox{$p\in priorities(e)$}
\end{align*}}
where $\textit{priorities}(e)$, resp., $\textit{unmatched-priorities}(e)$, is the set of priorities occurring in $\textit{put}$ operations of $e$, resp., in $\textit{put}$ operations of $e$ for which there is no $\textit{rm}$ operation removing the same value. We call the latter \emph{unmatched} put operations. A put operation which is not unmatched is called \emph{matched}. For simplicity, we consider the following syntactic sugar $\mathsf{Has\text{-}MatchedMaxPriority}(e)=\neg \mathsf{Has\text{-}EmptyRemoves}(e)\land \neg \mathsf{Has\text{-}UnmatchedMaxPriority}(e)$. By an abuse of notation, we also assume that  $\mathsf{Has\text{-}UnmatchedMaxPriority}(e) \Rightarrow \neg \mathsf{Has\text{-}EmptyRemoves}(e)$ (this is sound by the order of the conditionals in $\textit{Check-PQ-Seq}$).

The predicates defining the correctness of the selected operations are defined as follows:
{\small
\begin{align*}
\mathsf{EmptyRemove\text{-}Seq}(e,o)=\mathsf{true} & \mbox{ iff  $e= u\cdot o\cdot v$ and $\textit{matched}(u)$} \\
\mathsf{UnmatchedMaxPriority\text{-}Seq}(e,x)=\mathsf{true} & \mbox{ iff  $e= u\cdot \textit{put}(x,p)\cdot v$, $p\not\prec \textit{priorities}(u\cdot v)$, and $p\not\in \textit{priorities}(v)$} \\
\mathsf{MatchedMaxPriority\text{-}Seq}(e,x)=\mathsf{true} & \mbox{ iff  $e= u\cdot \textit{put}(x,p)\cdot v\cdot \textit{rm}(x)\cdot w$, $p\not\prec \textit{priorities}(u\cdot v\cdot w)$,} \\
&\hspace{4mm}\mbox{$p\not\preceq \textit{unmatched-priorities}(u\cdot v\cdot w)$, $\textit{matched}_\prec(u\cdot v,p)$,} \\
&\hspace{4mm}\mbox{and $p\not\in \textit{priorities}(v\cdot w)$}
\end{align*}
}

\vspace{-3mm}
\noindent
where $p\prec \textit{priorities}(e)$ when $p\prec p'$ for some $p'\in \textit{priorities}(e)$ (and similarly for $p\prec \textit{unmatched-p}$ $\textit{riorities}(e)$ or $p\preceq \textit{unmatched-priorities}(e)$),
$\textit{matched}_\prec(e,p)$ holds when each value with priority strictly smaller than $p$ is removed in $e$, and $\textit{matched}(e)$ holds when $\textit{matched}_\prec(e,p)$ holds for each $p\in\mathbb{P}$. Compared to the example presented at the beginning of the section, these predicates take into consideration that multiple values with the same priority are removed in FIFO order: the predicate $\mathsf{MatchedMaxPrioritySeq}(e,x)$ holds when $x$ is the last value with priority $p$ added in $e$.

When $o$ is a $\textit{rm}(\textit{empty})$ operation, we write $e\setminus o$ for the maximal subsequence of $e$ which doesn't contain $o$. For an execution $e$, $\textit{values}(e)$ is the set of values occurring in call/return actions of $e$.

% and items of unmatched put in $l$, respectively. $p \prec \textit{priorities}(l)$ (resp., $p \preceq \textit{priorities}(l)$), if $p \prec p'$ (resp., $p \preceq p'$)for some $p' \in \textit{priorities}(l)$. Similarly we can define $p \preceq \textit{unmatched-priorities}(l)$.
%
%
%Essentially, this procedure chooses either a $\textit{rm}(\textit{empty})$ operation or a set of operations adding/removing a maximal priority in the input execution and checks whether they satisfy certain properties
%
%
% proceeds by
%
% The projection $e \vert{O}$ of $e$ into $O$ is obtained from $e$ by erasing all call and return actions of non-$O$ operations. We write $e \setminus o$ for the projection $e \vert_{ O_e \setminus \{ o \} }$, where $O_e$ is the set of operations of $e$.
%
%Given a data-differentiated sequential execution $e$, it is costly to check whether $e$ belongs to priority queue, since for each element of $e$, we should keep the current content of priority queue and try to modify the priority queue according to this element. Such process is too complex. In this section, we propose another method for checking inclusion of priority queue by using a recursive procedure $\textit{check-PQ}$. For checking a sequence, every time we only check a much simpler property, and then recursively check the remanning sequences obtained by erasing one or two elements. Each step of our method is easier to deal with.
%
%$\textit{put}(a,p)$ matches $\textit{rm}(b)$, if $a = b$. To introduce $\textit{check-PQ}$, let us introduce several predicates. Given sequential execution $l$ and priority $p$:
%
%\begin{itemize}
%\setlength{\itemsep}{0.5pt}
%\item[-] $\textit{rm}(\textit{empty}) \notin l$ is satisfied when $l$ does not contain $\textit{rm}(\textit{empty})$.
%
%\item[-] Let $\textit{priorities}(l)$ and $\textit{unmatched-priorities}(l)$ be the set of priorities of putted items and items of unmatched put in $l$, respectively. $p \prec \textit{priorities}(l)$ (resp., $p \preceq \textit{priorities}(l)$), if $p \prec p'$ (resp., $p \preceq p'$)for some $p' \in \textit{priorities}(l)$. Similarly we can define $p \preceq \textit{unmatched-priorities}(l)$.
%
%\item[-] $\textit{matched}_{\prec}(l,p)$ is satisfied, if for each item of $l$ with priorities smaller than $p$, it has matched $\textit{put}$ and $\textit{rm}$. Similarly, we can define $\textit{matched}(l)$, where we consider all items in $l$ instead of items with certain priorities.
%\end{itemize}
%
%Let us introduce some notations: given sequential executions $u,v,w$ and priority $p$,
%
%\begin{itemize}
%\setlength{\itemsep}{0.5pt}
%\item[-] $\textit{PQ}_1(u,v,w,p) \equiv
%(\textit{rm}(\textit{empty}) \notin u \cdot v \cdot w) \wedge
%(p \not\prec \textit{priorities}(u \cdot v \cdot w)) \wedge
%(p \not\preceq \textit{unmatched-priorities}(u \cdot v \cdot w)) \wedge
%(\textit{matched}_{\prec}(u \cdot v,p) ) \wedge
%(p \notin \textit{priorities}(v \cdot w))$
%
%\item[-] $\textit{PQ}_2(u,v,p) \equiv
%(\textit{rm}(\textit{empty}) \notin u \cdot v) \wedge
%(p \not\prec \textit{priorities}(u \cdot v \cdot w)) \wedge
%(p \notin \textit{priorities}(v \cdot w))$
%
%\item[-] $\textit{PQ}_3(u,v) \equiv
%\textit{matched}(u \cdot v)$
%\end{itemize}
%
%
%Let us defined set $\textit{MS}(R)$ ($R \in \textit{PQ}_1,\textit{PQ}_2,\textit{PQ}_3$), which is the set of data-differentiated sequential executions that respect $R$.
%
%\begin{itemize}
%\setlength{\itemsep}{0.5pt}
%\item[-] $e \in \textit{MS}(\textit{PQ}_1)$, if $e = u \cdot \textit{put}(x,p) \cdot v \cdot \textit{rm}(x) \cdot w$ and $\textit{PQ}_1(u,v,w,p)$ holds for some item $x$ and a maximal priority $p$,
%
%\item[-] $e \in \textit{MS}(\textit{PQ}_2)$, if $e = u \cdot \textit{put}(x,p) \cdot v$ and $\textit{PQ}_2(u,v,p)$ holds for some item $x$ and a maximal priority $p$,
%
%\item[-] $e \in \textit{MS}(\textit{PQ}_3)$, if $e = u \cdot \textit{rm}(\textit{empty}) \cdot v$ and $\textit{PQ}_3(u,v)$ holds.
%\end{itemize}
%
%We call such $x$ and $\textit{rm}(\textit{empty})$ (if exists) the witness of $e$. Note that sequences in $\textit{MS}(R)$ does not guaranteed to be a priority queue execution. Let us introduce the notion $\textit{last}(e)$, which is a set and used as the branch condition of $\textit{check-PQ}$. Here $e$ is a data-differentiated sequential execution.
%
%\begin{itemize}
%\setlength{\itemsep}{0.5pt}
%\item[-] If $e$ contains $\textit{rm}(\textit{empty})$, then $\textit{last}(e) = \{ \textit{PQ}_3 \}$.
%
%\item[-] Else, if for a maximal priority $p$ of $e$, $\textit{unmatched-priorities}(e) \neq \emptyset$, then $\textit{PQ}_2 \in \textit{last}(e)$.
%
%\item[-] Else, if for a maximal priority $p$ of $e$, $\textit{unmatched-priorities}(e) = \emptyset$, then $\textit{PQ}_1 \in \textit{last}(e)$.
%\end{itemize}
%
%When $\textit{last}(e)$ contains only one element $R$, we write $\textit{last}(w)=R$ for simplicity.
%
%The projection $e \vert{\mathcal{D}}$ of $e$ into $\mathcal{D} \subseteq \mathbb{D}$ is obtained from $e$ by erasing all elements with a data value not in $\mathcal{D}$. We write $e \setminus x$ for the projection $e \vert_{ \mathbb{D} \setminus \{ x \} }$. The projection $e \vert{O}$ of $e$ into $O$ is obtained from $e$ by erasing all call and return actions of non-$O$ operations. We write $e \setminus o$ for the projection $e \vert_{ O_e \setminus \{ o \} }$, where $O_e$ is the set of operations of $e$.
%
%The recursive procedure $\textit{con-check-EPQ}$ is given as follows. Here $P_i(e,o)$ and $P_i(e,x)$ is the predicates of $e \in \textit{MS}(\textit{PQ}_i)$ withe witness $o$ or $x$, respectively.
%
%\begin{minipage}{.5\textwidth}
%\begin{algorithm}[H]
%\KwIn {a data-differentiated sequential execution $e$}
%
%\If {$e = \epsilon$}
%{\KwRet $\textit{true}$;}
%
%\If {$\textit{last}(e) = \textit{PQ}_3$}
%{
%    \If {$\exists o=\textit{rm}(\textit{empty})$, $P_3(e,o)$ holds}
%    {
%        \KwRet $\textit{check-PQ}(e \setminus o)$;
%    }
%    \KwRet $\textit{false}$;
%}
%
%\If {$\textit{PQ}_i \in \textit{last}(e)$ for $i \in \{1,2\}$}
%{
%    \If {$\exists l$, $P_i(e,x)$ holds}
%    {
%        \KwRet $\textit{check-PQ}(e \setminus x)$;
%    }
%    \KwRet $\textit{false}$;
%}
%\caption{$\textit{check-PQ}$}
%\label{Method-check-PQ}
%\end{algorithm}
%\end{minipage}
%\hfill

%TODO MODIFY THE FOLLOWING EXAMPLE
%
%Let us use an example to show how $\textit{check-PQ}$ works. Here we use $w_1 \xrightarrow{R} w_2$ to denote that in $\textit{check-PQ}(w_1)$, we choose the branch of $R$ and recursively call $\textit{check-PQ}(w_2)$.
%
%\begin{example}\label{example:generate extended priority queue executions}
%Given priorities $p_1,p_2,p_3$ with orders $p_1 \prec p_2$ and $p_1 \prec p_3$,
%
%\noindent $\textit{put}(c,p_2) \cdot \textit{put}(a,p_1) \cdot \textit{rm}(a) \cdot \textit{rm}(c) \cdot \textit{rm}(\textit{empty}) \cdot \textit{put}(d,p_2) \cdot \textit{put}(e,p_3) \cdot \textit{rm}(e) \cdot \textit{put}(b,p_1)$
%
%$\xrightarrow{\textit{EPQ}_3}$ $\textit{put}(c,p_2) \cdot \textit{put}(a,p_1) \cdot \textit{rm}(a) \cdot \textit{rm}(c) \cdot \textit{put}(d,p_2) \cdot \textit{put}(e,p_3) \cdot \textit{rm}(e) \cdot \textit{put}(b,p_1)$
%
%$\xrightarrow{\textit{EPQ}_1}$ $\textit{put}(c,p_2) \cdot \textit{put}(a,p_1) \cdot \textit{rm}(a) \cdot \textit{rm}(c) \cdot \textit{put}(d,p_2) \cdot \textit{put}(b,p_1)$
%
%$\xrightarrow{\textit{EPQ}_2}$ $\textit{put}(c,p_2) \cdot \textit{put}(a,p_1) \cdot \textit{rm}(a) \cdot \textit{rm}(c) \cdot \textit{put}(b,p_1)$
%
%$\xrightarrow{\textit{EPQ}_1}$ $\textit{put}(a,p_1) \cdot \textit{rm}(a) \cdot \textit{put}(b,p_1)$
%
%$\xrightarrow{\textit{EPQ}_2}$ $\textit{put}(a,p_1) \cdot \textit{rm}(a)$
%
%$\xrightarrow{\textit{EPQ}_1}$ $\epsilon$
%\end{example}

@@

\begin{restatable}{lemma}{EPQRulesAndSemantics}
\label{lemma:EPQ rules and semantics}
$\textit{Check-PQ-Seq}(e)=\mathsf{true}$ iff $e\in \seqPQ$, for every data-differentiated sequential execution $e$.
\end{restatable}

%
%\begin{lemma}
%Algorithm~\ref{alg:seq_check} returns $\mathsf{true}$ on an execution $e$ iff $e\in \seqPQ$.
%\end{lemma}

%Let $\textit{PQ}$ be set of sequences obtained by renaming some data-differentiated sequential execution that is accepted by $\textit{check-PQ}$. To persuade readers that $\textit{PQ}$ is indeed the set of behaviors of priority queue, in Appendix \ref{sec:appendix in section inductive rules of extended priority queue}, we give a semantical version definition $\textit{PQ}_s$ of priority queue, and shows that $\textit{PQ} = \textit{PQ}_s$. To model the possible behaviors of priority queue, we model it as an labelled transition system (shortened as LTS) $\textit{LTS}_e$. Each state of $\textit{LTS}_e$ is a function from $\mathbb{P}$ into sequences over $\mathbb{D}$, and represents a snapshot of contents of priority queue. $\textit{PQ}_s$ is the set of traces of $\textit{LTS}_e$. The definition of $\textit{PQ}_s$ and the proof of the following lemma can be found in Appendix \ref{sec:appendix in section inductive rules of extended priority queue}.
%
%\begin{restatable}{lemma}{EPQRulesAndSemantics}
%\label{lemma:EPQ rules and semantics}
%$\textit{PQ} = \textit{PQ}_s$.
%\end{restatable}
%
%By Example \ref{example:generate extended priority queue executions} we can see how to generate a sequence $e \in \textit{EP}$ from $\epsilon$ as follows:
%
%\begin{itemize}
%\setlength{\itemsep}{0.5pt}
%\item[-] First we add non-$\textit{rm}(\textit{empty})$ elements with a loop: Let $P$ be the set of priorities of items in $e$. In the first round of the loop, we choose a minimal priorities $p \in P$, we add items by using $\textit{PQ}_1$ and $\textit{PQ}_2$, let $P = P \setminus \{p\}$, and begins the next round of loop.
%
%\item[-] Then, we add $\textit{rm}(\textit{empty})$ by using $\textit{PQ}_3$.
%\end{itemize}

@@

The extension of $\textit{Check-PQ-Seq}$ to concurrent executions, checking whether they are linearizable w.r.t. $\seqPQ$, is obtained by replacing every predicate $\Gamma\mathsf{\text{-}Seq}$ with
\begin{align*}
\Gamma\mathsf{\text{-}Conc}(e,\alpha) = \mathsf{true}\mbox{ iff there exists a sequential execution $s$ such that $e\sqsubseteq s$ and $\Gamma\mathsf{\text{-}Seq}(s,\alpha)$}
\end{align*}
for each $\Gamma\in \{\mathsf{EmptyRemove}, \mathsf{UnmatchedMaxPriority}, \mathsf{MatchedMaxPriority}\}$. Let $\textit{Check-PQ-Conc}$ denote the thus obtained procedure (we assume recursive calls are modified accordingly).

% with linearizability to deal with linearizability w.r.t $\textit{PQ}$ as follows:
%
%\begin{itemize}
%\setlength{\itemsep}{0.5pt}
%\item[-] The input $e$ is a concurrent execution.
%
%\item[-] $\textit{last}(e)$ is the union of $\textit{last}(u)$ for any sequential execution $u$ such that $e \sqsubseteq u$. $e \sqsubseteq \textit{MS}(R)$ with witness $x$ (resp., $o$), if $e \sqsubseteq u \in \textit{MS}(R)$ and $x$ (resp., $o$) is the witness of $u$.
%
%\item[-] $P_i(e,o)$ and $P_i(e,x)$ holds, if $e \sqsubseteq \textit{MS}(\textit{PQ}_i)$ withe witness $o$ or $x$, respectively.
%\end{itemize}

The following lemma states the correctness of $\textit{Check-PQ-Conc}$. Completeness follows easily from the properties of $\seqPQ$. Thus, if $\textit{Check-PQ-Conc}(e) = \textit{false}$, then there exists a set $D$ of values s.t. either  $\mathsf{EmptyRemove\text{-}Conc}(e \vert D)$ is false, or $\mathsf{UnmatchedMaxPriority\text{-}Conc}(e \vert D,x)$ is false for all the values $x$ of maximal priority that are not removed (and there exists at least one such value), or $\mathsf{MatchedMaxPriority\text{-}Conc}(e \vert D,x)$ is false for all the values $x$ of maximal priority (and these values are all removed in $e \vert D$). It can be easily seen that we get $e \vert D\not\sqsubseteq \seqPQ$ in all cases, which by the closure under projection of $\seqPQ$ implies, $e \not\sqsubseteq \seqPQ$ (since every linearization of $e$ includes as a subsequence a linearization of $e \vert D$).

\begin{restatable}{lemma}{ConCheckEPQIsCorrect}
\label{lemma:con-check-EPQ is correct}
$\textit{Check-PQ-Conc}(e)=\mathsf{true}$ iff $e \sqsubseteq \seqPQ$, for every data-differentiated execution $e$.
\end{restatable}

>>>>

We define a recursive procedure for checking linearizability of a data-differentiated execution w.r.t. $\seqPQ$.
To ease the exposition, Section~\ref{ssec:seq_exec} introduces a recursive procedure for checking whether a data-differentiated \emph{sequential} execution is admitted by the priority queue which is then extended to the concurrent case in Section~\ref{ssec:conc_exec}.

@@

The recursive procedure $\textit{Check-PQ-Seq}$ outlined in Algorithm~\ref{alg:seq_check} checks whether a data-differentiated sequential execution belongs to $\seqPQ$ (i.e., if it is accepted by the LTS $PQ$).
 Roughly, it selects one or two operations in the input execution, checks whether their return values are correct by ignoring the order between the other operations other than how they are ordered w.r.t. the selected ones, and calls itself recursively on the execution without the selected operations.

We explain how the procedure works on the following execution:
\begin{align}
\hspace{-5mm}\textit{put}(c,p_2)\hspace{-.5mm} \cdot \textit{put}(a,p_1) \cdot \textit{rm}(a) \cdot \textit{rm}(c) \cdot \textit{rm}(\textit{empty}) \cdot \textit{put}(d,p_2) \cdot \textit{put}(\mathit{f},p_3) \cdot \textit{rm}(\mathit{f}) \cdot \textit{put}(b,p_1)\label{eq:ex_rec1}
\end{align}
where $p_1$, $p_2$, $p_3$ are priorities such that $p_1 \prec p_2$ and $p_1 \prec p_3$, and $p_2$ and $p_3$ are incomparable. Since the $\textit{rm}(\textit{empty})$ operations
are read-only (they don't affect the queue's state), they are selected first. An $\textit{rm}(\textit{empty})$-operation $o$ is correct when every $\textit{put}(x,p)$ operation before $o$ is matched to a $\textit{rm}(x)$ operation which also occurs before $o$. This is true in this case for $x\in \{a,c\}$. Thus, the correctness of (\ref{eq:ex_rec1}) reduces to the correctness of
\begin{align}
\textit{put}(c,p_2) \cdot \textit{put}(a,p_1) \cdot \textit{rm}(a) \cdot \textit{rm}(c) \cdot \textit{put}(d,p_2) \cdot \textit{put}(f,p_3) \cdot \textit{rm}(f) \cdot \textit{put}(b,p_1)\label{eq:ex_rec2}
\end{align}
When the execution contains no $\textit{rm}(\textit{empty})$-operation, the procedure selects a $\textit{put}$ operation adding a value that is not removed and that has a maximal priority. For (\ref{eq:ex_rec2}), it selects $\textit{put}(d,p_2)$ because $p_2$ is a maximal priority. This operation is correct since $d$ is the last value with priority $p_2$ in the execution, and the correctness of (\ref{eq:ex_rec2}) reduces to the correctness of
\begin{align}
\textit{put}(c,p_2) \cdot \textit{put}(a,p_1) \cdot \textit{rm}(a) \cdot \textit{rm}(c) \cdot \textit{put}(f,p_3) \cdot \textit{rm}(f) \cdot \textit{put}(b,p_1)\label{eq:ex_rec3}
\end{align}
If no operations like above can be found, $\textit{Check-PQ-Seq}$ selects a pair of $\textit{put}$ and $\textit{rm}$ operations adding and removing the same maximal priority value. For (\ref{eq:ex_rec2}), it can select
$\textit{put}(c,p_2)$ and $\textit{rm}(c)$. The value returned by $\textit{rm}(c)$ is correct if all the values of priority smaller than $p_2$ added before $\textit{rm}(c)$ are also removed before $\textit{rm}(c)$. In this case, $a$ is the only value of priority smaller than $p_2$ and it satisfies this property. Applying a similar reasoning for all the remaining values, it can be proved that this execution is correct.

\begin{algorithm}[t]
\footnotesize{
\KwIn {A data-differentiated sequential execution $e$}
\KwOut{$\mathsf{true}$ iff $e\in \seqPQ$}

@@

\If {$\mathsf{Has\text{-}EmptyRemoves}(e)$}
{
    \If {$\exists\ o=\textit{rm}(\textit{empty})\in e$ such that $\mathsf{EmptyRemove\text{-}Seq}(e,o)$ holds}
    {
        \KwRet $\textit{Check-PQ-Seq}(e \setminus o)$\;
    }
}
\ElseIf{$\mathsf{Has\text{-}UnmatchedMaxPriority}(e)$}
{
    \If {$\exists\ x \in \textit{values}(e)$ such that $\mathsf{UnmatchedMaxPriority\text{-}Seq}(e,x)$ holds}
    {
        \KwRet $\textit{Check-PQ-Seq}(e \setminus x)$\;
    }
}

\Else
{
    \If {$\exists\ x \in \textit{values}(e)$ such that $\mathsf{MatchedMaxPriority\text{-}Seq}(e,x)$ holds}
    {
        \KwRet $\textit{Check-PQ-Seq}(e \setminus x)$\;
    }
    \Else {\KwRet $\mathsf{false}$\;}
}}
\caption{$\textit{Check-PQ-Seq}$}
\label{alg:seq_check}
\end{algorithm}

Formally, the selected operations depend on the following set of predicates on executions:
\begin{align*}
& \mathsf{Has\text{-}EmptyRemoves}(e)=\mathsf{true} \mbox{ iff  $e$ contains a $\textit{rm}(\textit{empty})$-operation} \hspace{1cm}\\
& \mathsf{Has\text{-}UnmatchedMaxPriority}(e)=\mathsf{true} \mbox{ iff $p\in \textit{unmatched-priorities}(e)$ for a maximal $p$}
\end{align*}
where $\textit{priorities}(e)$, resp., $\textit{unmatched-priorities}(e)$, is the set of priorities occurring in $\textit{put}$ operations of $e$, resp., in $\textit{put}$ operations of $e$ for which there is no $\textit{rm}$ operation removing the same value. We call the latter \emph{unmatched} put operations. A put operation which is not unmatched is called \emph{matched}. For simplicity, we consider the following syntactic sugar $\mathsf{Has\text{-}MatchedMaxPriority}(e)=\neg \mathsf{Has\text{-}EmptyRemoves}(e)\land \neg \mathsf{Has\text{-}UnmatchedMaxPriority}(e)$. By an abuse of notation, we assume  $\mathsf{Has\text{-}UnmatchedMaxPriority}(e) \Rightarrow \neg \mathsf{Has\text{-}EmptyRemoves}(e)$ (this is sound by the order of the conditionals in $\textit{Check-PQ-Seq}$).

The predicates defining the correctness of the selected operations are defined as follows:
\begin{align*}
&\mathsf{EmptyRemove\text{-}Seq}(e,o)=\mathsf{true} \mbox{ iff  $e= u\cdot o\cdot v$ and $\textit{matched}(u)$} \\
&\mathsf{UnmatchedMaxPriority\text{-}Seq}(e,x)=\mathsf{true}  \mbox{ iff  $e= u\cdot \textit{put}(x,p)\cdot v$, $p\not\prec \textit{priorities}(u\cdot v)$,} \\
&\hspace{6.4cm}p\not\in \textit{priorities}(v) \\
&\mathsf{MatchedMaxPriority\text{-}Seq}(e,x)=\mathsf{true} \mbox{ iff  $e= u\cdot \textit{put}(x,p)\cdot v\cdot \textit{rm}(x)\cdot w$, $\textit{matched}_\prec(u\cdot v,p)$,} \\
&\hspace{4.3cm}\mbox{$p\not\preceq \textit{unmatched-priorities}(u\cdot v\cdot w)$, $p\not\prec \textit{priorities}(u\cdot v\cdot w)$,} \\
&\hspace{4.3cm}\mbox{and  $p\not\in \textit{priorities}(v\cdot w)$}
\end{align*}
where $p\prec \textit{priorities}(e)$ when $p\prec p'$ for some $p'\in \textit{priorities}(e)$ (and similarly for $p\prec \textit{unmatched-priorities}(e)$ or $p\preceq \textit{unmatched-priorities}(e)$),
$\textit{matched}_\prec(e,p)$ holds when each value with priority strictly smaller than $p$ is removed in $e$, and $\textit{matched}(e)$ holds when $\textit{matched}_\prec(e,p)$ holds for each $p\in\mathbb{P}$. Compared to the example presented at the beginning of the section, these predicates take into consideration that multiple values with the same priority are removed in FIFO order: the predicate $\mathsf{MatchedMaxPrioritySeq}(e,x)$ holds when $x$ is the last value with priority $p$ added in $e$.

When $o$ is an $\textit{rm}(\textit{empty})$-operation, $e\setminus o$ is the maximal subsequence of $e$ which doesn't contain $o$. For an execution $e$, $\textit{values}(e)$ is the set of values  in call/return actions of $e$.

@@

\begin{lemma}
\label{lemma:EPQ rules and semantics}
$\textit{Check-PQ-Seq}(e)=\mathsf{true}$ iff $e\in \seqPQ$, for every data-differentiated sequential execution $e$.
\end{lemma}

@@

The extension of $\textit{Check-PQ-Seq}$ to concurrent executions, checking whether they are linearizable w.r.t. $\seqPQ$, is obtained by replacing every predicate $\Gamma\mathsf{\text{-}Seq}$ with
\begin{align*}
\Gamma\mathsf{\text{-}Conc}(e,\alpha) = \mathsf{true}\mbox{ iff there is a sequential execution $s$ such that $e\sqsubseteq s$ and $\Gamma\mathsf{\text{-}Seq}(s,\alpha)$}
\end{align*}
for each $\Gamma\in \{\mathsf{EmptyRemove}, \mathsf{UnmatchedMaxPriority}, \mathsf{MatchedMaxPriority}\}$. The obtained procedure is denoted by
$\textit{Check-PQ-Conc}$ (recursive calls are modified accordingly).

The following lemma states the correctness of $\textit{Check-PQ-Conc}$. Completeness follows easily from the properties of $\seqPQ$. If $\textit{Check-PQ-Conc}(e) = \textit{false}$, then there exists a set $D$ of values s.t. either  $\mathsf{EmptyRemove\text{-}Conc}(e \vert D)$ is false, or $\mathsf{UnmatchedMaxPriority\text{-}Conc}(e \vert D,x)$ is false for all the values $x$ of maximal priority that are not removed (and there exists at least one such value), or $\mathsf{MatchedMaxPriority\text{-}Conc}(e \vert D,x)$ is false for all the values $x$ of maximal priority (and these values are all removed in $e \vert D$). It can be easily seen that we get $e \vert D\not\sqsubseteq \seqPQ$ in all cases, which by the closure under projection of $\seqPQ$ implies, $e \not\sqsubseteq \seqPQ$ (since every linearization of $e$ includes as a subsequence a linearization of $e \vert D$).

\begin{lemma}\label{lemma:con-check-EPQ is correct}
$\textit{Check-PQ-Conc}(e)=\mathsf{true}$ iff $e \sqsubseteq \seqPQ$, for every data-differentiated $e$.
\end{lemma}

Proving soundness is highly non-trivial and one of the main technical contributions of this paper (see Appendix~\ref{sec:appendix subsection proof of lemma con-check-EPQ is correct} for a complete proof). The main technical difficulty is showing that for any execution $e$, any linearization of $e\setminus x$ for some maximal priority value $x$ can be extended to a linearization of $e$ provided that $\mathsf{UnmatchedMaxPriority}$ or $\mathsf{MatchedMaxPriority}$ holds (depending on whether there are values with the same priority as $x$ in $e$ which are not removed).
